Matched Source Waveform Inversion for Transmitted Waves

### 2024-05-26 repro/papers/awinum/paper.tex
interim

\begin{abstract}
Matched Source Waveform Inversion applied to transmission data
produces an estimate of refractive index similar to the result of
travel time inversion, but without explicit identification of travel
times. This paper reviews the theoretical justification of this result
and its limitations, and exhibits a numerical illustration. 
\end{abstract}
\setlength{\parindent}{0cm}

\section{Introduction}
Matched Source Waveform Inversion (MSWI) is a variant of Full Waveform
Inversion (FWI) \cite[]{VirieuxOperto:09}, that sometimes overcomes
one of FWI's impediments, namely its tendency to stagnate at
suboptimal model estimates (``cycle-skip''), demonstrably far from global optima in
controlled settings and often uninformative of material structure in
field use. It is possible that such local descent algorithms are
trapped in regions around local minima, though such trapping, or even
the existance of local, non-global minima, are seldom established with
any rigor. MSWI
loosens the bond between predicted and observed data by interposing a
filter, adapted to map one to the other trace-by-trace, and penalizes
deviation of the filter from the Dirac delta. This soft penalty
effectively relaxes the FWI data-fitting problem and permits local
optimization methods to approximate data kinematics.

The aim of this paper is to review the theoretical basis of MSWI,
describe a practical computational framework for MSWI, and use it to
solve a synthetic acoustic inversion problem for which FWI fails. In
contrast, MSWI starting at the same initial estimate delivers a
improved velocity from which FWI succeeds in approaching the global
minimizer. The example is representative of a case (single arrival or
simple wavefront transmitted wave data) in which the MSWI objective
function is demonstrably close to mean square travel time error. Since
travel time inversion is generally not subject to cycle-skipping,
successful inversion is expected, and that is what we observe.

@@

\section{Background}
The version of MSWI discussed here uses acoustic wave propagation with
isotropic point sources and receivers.
The pressure and velocity fields $p({\bf x},t;{\bf x}_s)$, ${\bf v}({\bf x},t;{\bf x}_s)$ for the source location ${\bf x}_s$ depend on the bulk modulus $\kappa({\bf x})$, buoyancy $\beta({\bf x})$ (reciprocal of the density $\rho({\bf x})$), and wavelet $w(t;{\bf x}_s)$ through the acoustic system
\begin{eqnarray}
  \label{eqn:awe}
 \frac{\partial p}{\partial t} & = &- \kappa \nabla \cdot {\bf v} +
w(t;{\bf x}_s) \delta({\bf x}-{\bf x}_s); \nonumber \\
\frac{\partial {\bf v}}{\partial t} & = & - \beta \nabla p; \\ 
p, {\bf v} & = & 0 \mbox{ for }  t \ll 0.
\end{eqnarray}
The model vectors $m=(\kappa,\rho)$ make up the domain of the forward
map or {\em modeling operator} is $F[m]w = \{p({\bf x}_r,t;{\bf
  x}_s)\}$, for specified source and receiver positions ${\bf x}_s, {\bf x}_r$ and
recording interval $[0,t_d]$.

@@

The approach to local optimization taken here (and in most work on FWI
and related topics) is based on the gradient of the objective defined
in equiation \ref{eqn:fwi}:
\begin{equation}
  \label{eqn:fwigrad}
  g = \nabla  J_{\rm FWI}[m;d] = D_m(F[m]w)^T(F[m]w-d).
\end{equation}
In this formula, $D_m(F[m]w)$ is the derivative of $F[m]w$ with
respect to $m$. 
This is the Euclidean (or $L^2$) gradient, that is, the vector $g$ for
which the Euclidean inner product
\begin{equation}
  \label{eqn:eucip}
  \langle g, \delta m\rangle = g^T\delta m
\end{equation}
with any other vector $\delta m$ gives the
rate of change in the direction of that vector of $J_{\rm FWI}$ at m.

@@

Note that the filtered predicted data may be viewed as the predicted
data for an {\em extended source}, with the source wavelet at location
$\bx_s$ replaced by $u(\bx_s,\bx_r,\cdot) * w(\bx_s,\cdot)$. That is,
the adaptive filter construction is equivalent to allowing the source
to depend on receiver position as well as source position - one source
wavelet for each source-receiver pair. This is an extension of
standard modeling, in that the domain (bulk modulus buoyancy, wavelet)
is larger than in the conventional formulation, and coincides with it
under the special condition that the source wavelet is independent of
the receiver location. This  {\em
  source-receiver extension}\cite[]{HuangSymes2015SEG} is a key ingredient in a number of
other papers on modifications of FWI, as mentioned in the Introduction.

@@

\cite{Symes:24a} explains the relation between the reduced MSWI
objective and travel time inversion. This relation follows from the
geometric asymptotics approximation to solutions of the point radiator
problem \ref{eqn:awe} \cite[]{Friedlander:75}. Presuming that the
coefficients $\kappa, \beta$ are smooth, there are smooth functions
$\tau[m](\bx_s,\bx)$ (travel time) and $a[m](\bx_s,\bx)$ (geometric
amplitude) depending on the model $m$, so that
\begin{equation}
  \label{eqn:pwa}
  p(\bx,t;\bx_s) = a[m](\bx,\bx_s)w(t-\tau[m](\bx,\bx_s),\bx_s) + ...
\end{equation}
The elided terms are smoother than $w$, and become small with the
dominant wavelength in $w$ (\cite{Symes:94a} explains the precise
meaning of this condition). Thus we can write
\[
  F[m]w(\bx_r,t;\bx_s) \approx
  a[m](\bx_r,\bx_s)w(t-\tau[m](\bx_r,\bx_s))
\]
Supposing that the data $d$ is noise-free,
\[
  d(\bx_r,t;\bx_s) = F[m^*]w(\bx_r,t;\bx_s) \approx
  a[m^*](\bx_r,\bx_s)w(t-\tau[m^*](\bx_r,\bx_s))
\]
it follows that
\begin{equation}
  \label{eqn:tomo}
  \lim_{\alpha \rightarrow 0}\frac{1}{\alpha^2} (\tilde{J}_{\alpha,\sigma}[m;d]
  -\tilde{J}_{0,\sigma}) \approx \sum_{\bx_s,\bx_r} \frac{a[m^*]^2}{a[m]^2} (\tau[m^*]-\tau[m])^2\|g_{\frac{a[m^*]}{a[m]}\sigma}\|^2
\end{equation}
Here $g_{\sigma}$ is an approximate delta. Its Fourier transform of
is
\begin{equation}
  \label{eqn:gsig}
  \hat{g}_{\sigma} = \frac{a[m^*]^2 |\hat{w}|^2}{a[m^*]^2 |\hat{w}|^2
    + \sigma^2}
\end{equation}
whence $g_{\sigma}(\bx_s,\bx_r,t) \rightarrow \delta(t)$ in
the sense of distributions as $\sigma \rightarrow 0$.

If $\alpha$ and $\sigma$ are sufficiently small, then
$\tilde{J}_{\alpha,\sigma}$ is close to the right-hand side, which is bounded above and below by multiples of the the mean square travel time error between $m$ and
$m^*$.  This is the relation mentioned in the introduction. 
It does not
show that the only stationary points are those of the mean square
travel time error, even approximately. However if the amplitudes are
relatively insensive to changes in model, as is true if the
source-receiver distance is well away from developing multiple
arrivals, then the difference of stationary points gets small as $m$
approaches a stationary point of $\tilde{J}_{\alpha,\sigma}$. So we
would expect minimization $\tilde{J}_{\alpha,\sigma}$ to produce a
model that closely matches the travel times inherent in the data
($\tau[m^*]$ in the notation used here).

AWI adds one more feature, namely scaling by the trace norm of the
filter. This normalization can be interpreted as a choice of weighted
norm on the space of adaptive filters. It leads to a very similar
relation to \ref{eqn:tomo} for the AWI penalty function, but without
the amplitude factors: that is, the right-hand side of \ref{eqn:tomo}
is {\em just} the mean-square traveltime error. See \cite{Symes:24a}
for details.

@@

\section{A numerical illustration}

@@

This section presents the inversion of acoustic data with dimensions
typical of crustal seismic exploration. The target bulk modulus field ($\kappa$)
is depicted in Figure \ref{fig:m}, and contains an acoustic ``lens''
positioned in the center between 1000 m and 3000 m depth
(``$z$''). The background level outside the lens is 4 GPa; at the
center it is 2 GPa. The
buoyancy ($\beta$)is spatially homogeneous at 1 cm$^3/g$, so the background
wave speed is 2000 m/s.

@@

This field is sampled on a 20 $\times$ 20 m grid. Acoustic wave
propagation is simulated via a staggered grid finite difference method
\cite[]{vir86,lev88,Cohen:01}. The time
step is chosen to be safely stable, given the parameter fields,
difference formulae, and
spatial sampling.

Receivers are spaced on the line at depth 1000 m, from 2000 m to 6000
m horizontal coordinate (``$x$''), spaced 20 m apart. Eleven source
positions lie on the line $z=3000$ m, spaced 200 m apart, from
$x=3000$ m to $x = 5000$ m. The source wavelet $w$ (the same for every
source ) is a $[1.0, 2.5, 7.5, 12.5]$ Hz trapezoidal bandpass filter,
with a median frequency of 5.875 Hz corresponding to a median wavelength of
$\approx$ 340 m, and shortest wavelength of 160 m. It is centered at
$t=1$ s.

The discrete pressure field is
sampled at the externally specified time grid and spatial locations
via piecewise linear interpolation in space and cubic spline
interpolation in time. The isotropic point source is added
into the acoustic fields at each time step via the adjoints of these
interpolation operations.

Simulated data from this configuration appears as Figure \ref{fig:d11}.

The initial model $m_0$ for inversion is homogeneous, with $\kappa = $ 4 GPa,
$\beta = $ 1 cm$^3$/g. The corresponding data is depicted in Figure \ref{fig:sim0}.
These plots show the time shift between the homogeneous model data and
the target data, for the central source positions.

\multiplot{2}{d11,sim0}{width=0.45\textwidth}{a: Data for target model $m$
  (Figure \ref{fig:m}). b: Data for homogeneous model $m_0$.}

We apply a version of weighted steepest descent optimization to the
minimization of $J_{\rm FWI}$. The inverse weight operator ($W^{-1}$
in formula \ref{eqn:fwiwgrad}) is a
10-point moving average in both spatial directions, repeated once. As
noted above, the weight operator itself is not required. The weighted
gradient is computed via formula \ref{eqn:fwigrad}. The adjoint
derivative $DF[m]^T$ is computed via the {\em adjoint state method}
\cite[]{Chavent:74,GauTarVir:86}, with time reversal of the acoustic fields
implemented via optimal checkpointing
\cite[]{Griewank:92,Griewank:book,Symes:06a-pub}. The negative of the weighted
gradient ($g_W$, formula \ref{eqn:fwiwgrad}) is the search direction,
and the optimal step in this direction is approximated by a simple
backtracking line search algorithm \cite[]{NocedalWright}.

The value of $J_{\rm FWI}[m_0,d] \approx 3.04$, and the rate of
increase in the weighted gradient direction at $m_0$ is 1.74 $\times
10^{-6}$. Note that the rate of increase is the same as the weighted
gradient norm. After 10 steepest descent steps, the objective value has
decreased by about 30\%, to 1.9, and the weighted gradient norm by
more than an order of magnitude, to $\approx 9.6 \times 10^{-8}$. The
final model appears in Figure \ref{fig:mestfwi0}, and the
corresponding data in Figure \ref{fig:resimfwi0}.

@@

\plot{mestfwi0}{width=\textwidth}{a: Bulk modulus produced by 10
  steepest descent steps to minimize the FWI objective \ref{eqn:fwi},
  using the data shown in \ref{fig:d11}.}

\multiplot{2}{resimfwi0,residfwi0}{width=0.45\textwidth}{a: Data
  corresponding to FWI inversion result shown in Figure
  \ref{fig:mestfwi0}. b: Residual or data error - difference between
  data shown in Figure \ref{fig:resimfwi0} and target data
  \ref{fig:d11}. Note the failure to match the later signal in the
  central part of the display.}

@@

Now apply the same algorithm to the MSWI objective
$J_{\alpha,\sigma}$. Of course, to do so requires a choice of the
parameters $\alpha$ and $\sigma$. The purpose of the regularization
parameter $\sigma > 0$ is
simply to assure that the system \ref{eqn:normal} is positive
definite, so it can be chosen quite small - whatever that means. The
penalty parameter $\alpha > 0$ is the link between the filtered
residual and the spread in time of the filter: the performance of the
algorithm depends primarily on its choice. The next section will
discuss rational approaches to this choice. For now, we use the values
$\alpha=0.0001, \sigma=0.00001$ based on numerical exploration of the
objective.

The system \ref{eqn:normal} is far larger than is convenient to solve
by any variant of Gaussian Elimination, so iterative methods are 
necessary. We choose the Conjugate Gradient (CG.) method
\cite[]{Dan:67,Steihaug:83,NocedalWright}, and stop the
iteration by monitoring the reduction in length of the normal residual
(the gradient of $J_{\alpha,\sigma}[m,u;d]$ with respect to $u$). The
reduction threshhold $\rho$ is thus another necessary input parameter:
we choose $\rho = 0.001$ (note that unlike $\alpha$ and $\sigma$,
$\rho$ is dimensionless). We use the same inverse weight operator as
in the FWI example, namely a 10 $\times$ 10 moving average, repeated
once.

The initial bulk modulus field is once again homogeneous at 4 GPa. The
initial value of the reduced MSWI objective is $\approx 1.32 \times 10^{-2}$,
and the length of the (weighted) gradient is $\approx 6.37 \times
10^{-10}$. After 7 iterations, the normal residual reduction criterion
is satisfied. The value has decreased to $\approx 9.2 \times 10^{-4}$,
or a bit more than an order of magnitude reduction. The gradient
length is $\approx 4.33 \times 10^{-12}$, a reduction of more than two
orders of magnitude. The resulting bulk modulus is shown in Figure \ref{fig:mest11}.

\plot{mest11}{width=\textwidth}{a: Bulk modulus produced by 7
  steepest descent steps to minimize the reduced MSWI objective \ref{eqn:redfiltpen},
  using the data shown in \ref{fig:d11}.}

\multiplot{2}{sim11,resid11}{width=0.45\textwidth}{a: Data
  corresponding to MSWI result shown in Figure
  \ref{fig:mest11}. b: Residual or data error - difference between
  data shown in Figure \ref{fig:sim11} and target data
  \ref{fig:d11}. Note that the later signal in the central portion is
  still shifted, but by much less than in Figure \ref{fig:residfwi0}.}

Inspection of the residual plot \ref{fig:resid11} suggests that the
data predicted from the MSWI result is in many places within a
half-wavelength travel time shift of the data in Figure \ref{fig:d11}.
A half-wavelength is the conventional threshhold for FWI initial fit
error: below that it works, above that it doesn't. Thus it is tempting
to attempt FWI from using the MSWI result as initial estimate.

We applied 11 iterations of steepest descent to the FWI objective,
with initial estimate of $m$ equal to the final estimate from the MSWI
inversion just described (Figure \ref{fig:mest11}). The initial FWI
objective value was $\approx 4.3$, and the weighted gradient norm was
$\approx 2.15 \times 10^{-5}$.  After 11 steps of steepest descent,
the prescribed reduction ($10^{-2}$) of the weighted gradient norm was achieved,
and the objective value had decreased to $\approx 4.41 \times
10^{-2}$. The resulting bulk modulus field is depicted in Figure
\ref{fig:mestfwi}. The predicted data at this model, and the residual
data, are shown in Figures \ref{fig:resimfwi} and \ref{fig:residfwi}
respectively.

The FWI iteration starting at the MSWI result has effectively achieved
the global minimum of the FWI objective function: the data match is
very close. This type of result could perhaps have been achieved by
systematically increasing $\alpha$, as will be discussed in the next
section. However, the current result has a remarkable corollary,
provided that we are willing to assume that the intermediate values of
the FWI objective between steepest descent iterates are monotonically
decreasing. This condition could in principle be checked
computationally. However, for the moment let us accept that it is
so. Then we can conclude that the homogeneous model at which the first
FWI inversion was started must lie in a basin of attraction of a
non-global minimizer, since the value of the objective function at the
homogeneous model is actually {\em smaller} than that at the result of
MSWI inversion. However the descent sequence starting at that initial
estimate arrives at essentially zero residual. No descent path from
the homogeneous model can

>>>>

\begin{abstract}
Matched Source Waveform Inversion applied to acoustic transmission data
produces an estimate of refractive index similar to the result of
travel time inversion, but without explicit identification of travel
times. This paper reviews the theoretical justification of this result
and its limitations, and with 2D numerical illustrations. 
\end{abstract}
\setlength{\parindent}{0cm}

\section{Introduction}
Matched Source Waveform Inversion (MSWI) is a variant of Full Waveform
Inversion (FWI) \cite[]{VirieuxOperto:09}, that sometimes overcomes
one of FWI's impediments, namely its tendency to stagnate at
suboptimal model estimates (``cycle-skip''), demonstrably far from global optima in
controlled settings and often uninformative of material structure in
field use. It is possible that such local descent algorithms are
trapped in regions around local minima, though such trapping, or even
the existence of local, non-global minima, are seldom established with
any rigor. MSWI
loosens the bond between predicted and observed data by interposing a
filter, adapted to map one to the other trace-by-trace, and penalizes
deviation of the filter from the Dirac delta. This soft penalty
effectively relaxes the FWI data-fitting problem and permits local
optimization methods to approximate data kinematics.

The aim of this paper is to review the theoretical basis of MSWI,
describe a practical computational framework for MSWI, and use it to
solve a synthetic acoustic inversion problem for which FWI fails. In
contrast, MSWI starting at the same initial estimate delivers an
improved velocity from which FWI succeeds in approaching the global
minimizer. The example is representative of a case (single arrival or
simple wavefront transmitted wave data) in which the MSWI objective
function is demonstrably close to mean square travel time error. Since
travel time inversion is generally not subject to cycle-skipping,
successful inversion is expected, and that is what we observe.

@@

\section{Theory}
The version of MSWI discussed here uses acoustic wave propagation with
isotropic point sources and receivers.
The pressure and velocity fields $p({\bf x},t;{\bf x}_s)$, ${\bf v}({\bf x},t;{\bf x}_s)$ for the source location ${\bf x}_s$ depend on the bulk modulus $\kappa({\bf x})$, buoyancy $\beta({\bf x})$ (reciprocal of the density $\rho({\bf x})$), and wavelet $w(t;{\bf x}_s)$ through the acoustic system
\begin{eqnarray}
  \label{eqn:awe}
 \frac{\partial p}{\partial t} & = &- \kappa \nabla \cdot {\bf v} +
w(t;{\bf x}_s) \delta({\bf x}-{\bf x}_s); \nonumber \\
\frac{\partial {\bf v}}{\partial t} & = & - \beta \nabla p; \\ 
p, {\bf v} & = & 0 \mbox{ for }  t \ll 0.
\end{eqnarray}
The model vectors $m=(\kappa,\rho)$ make up the domain of the forward
map or {\em modeling operator} is $F[m]w = \{p({\bf x}_r,t;{\bf
  x}_s)\}$, for specified source and receiver positions ${\bf x}_s, {\bf x}_r$ and
recording interval $[0,t_d]$.

@@

The approach to local optimization taken here (and in most work on FWI
and related topics) is based on the gradient of the objective defined
in equiation \ref{eqn:fwi}:
\begin{equation}
  \label{eqn:fwigrad}
  g = \nabla  J_{\rm FWI}[m;d] = D_m(F[m]w)^T(F[m]w-d).
\end{equation}
In this formula, $D_m(F[m]w)$ is the derivative of $F[m]w$ with
respect to $m$. 
This is the Euclidean (or $L^2$) gradient, that is, the vector $g$ for
which the Euclidean inner product
\begin{equation}
  \label{eqn:eucip}
  \langle g, \delta m\rangle = g^T\delta m
\end{equation}
with any other vector $\delta m$ gives the
rate of change in the direction of that vector of $J_{\rm FWI}$ at $m$.

@@

Note that the filtered predicted data may be viewed as the predicted
data for an {\em extended source}, with the source wavelet at location
$\bx_s$ replaced by $u(\bx_s,\bx_r,\cdot) * w(\bx_s,\cdot)$. That is,
the adaptive filter construction is equivalent to allowing the source
to depend on receiver position as well as source position - one source
wavelet for each source-receiver pair. This is an extension of
standard modeling, in that the domain (bulk modulus buoyancy, wavelet)
is larger than in the conventional formulation, and coincides with it
under the special condition that the source wavelet is independent of
the receiver location. This  {\em
  source-receiver extension} \cite[]{HuangSymes2015SEG} is a key ingredient in a number of
other papers on modifications of FWI, as mentioned in the Introduction.

@@

\cite{Symes:24a} explains the relation between the reduced MSWI
objective and travel time inversion. This relation follows from the
geometric asymptotics approximation to solutions of the point radiator
problem \ref{eqn:awe} \cite[]{Friedlander:75}. Presuming that the
coefficients $\kappa, \beta$ are smooth, there are smooth functions
$\tau[m](\bx_s,\bx)$ (travel time) and $a[m](\bx_s,\bx)$ (geometric
amplitude) depending on the model $m$, so that
\begin{equation}
  \label{eqn:pwa}
  p(\bx,t;\bx_s) = a[m](\bx,\bx_s)w(t-\tau[m](\bx,\bx_s),\bx_s) + ...
\end{equation}
The elided terms are smoother than $w$, and become small with the
dominant wavelength in $w$ (\cite{Symes:24a} explains the precise
meaning of this condition). Thus we can write
\[
  F[m]w(\bx_r,t;\bx_s) \approx
  a[m](\bx_r,\bx_s)w(t-\tau[m](\bx_r,\bx_s))
\]
Supposing that the data $d$ is noise-free,
\[
  d(\bx_r,t;\bx_s) = F[m^*]w(\bx_r,t;\bx_s) \approx
  a[m^*](\bx_r,\bx_s)w(t-\tau[m^*](\bx_r,\bx_s))
\]
it follows that
\begin{equation}
  \label{eqn:tomo}
  \lim_{\alpha \rightarrow 0}\frac{1}{\alpha^2} (\tilde{J}_{\alpha,\sigma}[m;d]
  -\tilde{J}_{0,\sigma}) \approx \sum_{\bx_s,\bx_r} \frac{a[m^*]^2}{a[m]^2} (\tau[m^*]-\tau[m])^2\|g_{\frac{a[m^*]}{a[m]}\sigma}\|^2
\end{equation}
Here $g_{\sigma}$ is an approximate delta. Its Fourier transform of
is
\begin{equation}
  \label{eqn:gsig}
  \hat{g}_{\sigma} = \frac{a[m^*]^2 |\hat{w}|^2}{a[m^*]^2 |\hat{w}|^2
    + \sigma^2}
\end{equation}
whence $g_{\sigma}(\bx_s,\bx_r,t) \rightarrow \delta(t)$ in
the sense of distributions as $\sigma \rightarrow 0$.

If $\alpha$ and $\sigma$ are sufficiently small, then
$\tilde{J}_{\alpha,\sigma}$ is close to the right-hand side, which is bounded above and below by multiples of the the mean square travel time error between $m$ and
$m^*$.  This is the relation mentioned in the introduction. 
It does not
show that the only stationary points are those of the mean square
travel time error, even approximately. However if the amplitudes are
relatively insensive to changes in model, as is true if the
source-receiver distance is well away from developing multiple
arrivals, then any stationary point of $\tilde{J}_{\alpha,\sigma}$ is
either close to a tomographic stationary point, or far from $m^*$. So we
would expect minimization $\tilde{J}_{\alpha,\sigma}$ to produce a
model that closely matches the travel times inherent in the data
($\tau[m^*]$ in the notation used here).

AWI adds one more feature, namely scaling by the trace norm of the
filter. This normalization can be interpreted as a choice of weighted
norm on the space of adaptive filters. It leads to a very similar
relation to \ref{eqn:tomo} for the AWI penalty function, but without
the amplitude factors: that is, the right-hand side of the AWI
analogue of \ref{eqn:tomo}
is {\em just} the mean-square traveltime error, up to controllable
error, for single-arrival transmission data. See \cite{Symes:24a}
for details.

\cite{Symes:24a} also shows that if energetic multiple arrivals are
present in the data, for any reason, then cross-talk between travel
time branches destroys the relation between MSWI (or AWI) and any
version of the mean-square travel time error. The next sections of
this paper will illustrate the
success of MSWI for single-arrival data, and its failure for data with
multiple arrivals.

We end this section by overviewing the computations required for
minimization of $\tilde{J}_{\alpha,\sigma}$.

@@

\section{Numerical Illustration}

@@

This section presents application of MSWI to simulated acoustic data with dimensions
typical of crustal seismic exploration. We use several configurations
to illustrate the capabilities and limitations of this approach. In
all cases, we use MSWI to generate an initial bulk modulus estimate
input to FWI.

\subsection{Experimental setup}

The several experiments to be described below share a number of
features, set out in this section.

Acoustic wave
propagation is simulated via a staggered grid finite difference method
\cite[]{vir86,lev88,Cohen:01} of order 2 in time and 8 in space. The time
step is an internal detail of the simulation, chosen to be safely
stable, given the parameter fields, difference formulae, and 
spatial sampling. The discrete pressure field is
output over the time range $0 < t < 5$ s with sample interval $0.008$
s at externally specified receiver locations
via piecewise linear interpolation in space and cubic spline
interpolation in time. Isotropic point sources are added
into the acoustic fields at each time step via the adjoints of these
interpolation operations. These and other details of the simulator are
discussed in \cite{GeoPros:11}.

Spatial fields (bulk modulus, buoyancy, gradients) occupy a
rectangular region of 8 km horizontal (``x'') extent and 4 km vertical
(``z'') extent, sampled on a 20 $\times$ 20 m grid. For all except the
third example, receivers are located on the vertical line at
horizontal coordinate $x$= 5000 m, spaced 20 m apart, starting at
depth $z$ = 200 m. Twenty isotropic point sources occupy positions lie
on the line $x=3000$ m, spaced 150 m apart, starting at depth = 500
m. For the third example, source and receiver lines are moved 2000 m
further apart, to $x=2000$ m and $x=6000$ m respectively, other
parameters remaining the same. [Note that this geometry is a 2D
cartoon of that used in cross-well seismic data acquisition.] The
source wavelet $w$ (the same for every source ) is a
$[1.0, 2.5, 7.5, 12.5]$ Hz trapezoidal bandpass filter, with a median
frequency of 5.875 Hz corresponding to a median wavelength of
$\approx$ 340 m, and shortest wavelength of 160 m. It is centered at
$t=1$ s.

The initial model $m_0$ for MSWI is homogeneous, with $\kappa = $ 4
GPa, $\beta = $ 1 cm$^3$/g. As noted earlier, we regard $\beta$ as
fixed in this series of experiments, so mention of it will be
suppressed. The corresponding data for isotropic point sources at $x=3000$ m,
point receivers at $x=5000$ m, and the bandpass filter time dependence
described earlier, is depicted in
Figure \ref{fig:chwd20}. We use both $m_0$ and the output of MSWI as
initial models for FWI..

\plot{chwd20}{width=\textwidth}{Data for homogeneous model $m_0$: 20
  sources at $x=3000$ m, 201 receivers at $x=5000$ m.}

We apply a version of weighted steepest descent optimization to the
minimization of both $J_{\rm FWI}$ and $\tilde{J}_{\alpha,\sigma}$.
The inverse weight operator ($W^{-1}$
in formula \ref{eqn:fwiwgrad}) is a
10-point moving average in both spatial directions, repeated once,
except in the fourth example as discussed below. Recall that the weight operator itself is not required. The weighted
gradient is computed via formula \ref{eqn:fwigrad}. The adjoint
derivative $DF[m]^T$ is computed via the {\em adjoint state method}
\cite[]{Chavent:74,GauTarVir:86}, with time reversal of the acoustic fields
implemented via optimal checkpointing
\cite[]{Griewank:92,Griewank:book,Symes:06a-pub}. The negative of the weighted
gradient ($g_W$, formula \ref{eqn:fwiwgrad}) is the search direction,
and the optimal step in this direction is approximated by a simple
backtracking line search algorithm \cite[]{NocedalWright}. The
iteration is terminate either when the gradient norm has fallen below
1\% of its initial value, or when a maximum number of iterations
(usually 12) is reached.

The MSWI objective $\tilde{J}_{\alpha,\sigma}$ requires choices of the
parameters $\alpha$ and $\sigma$, and solution of the normal equation
\ref{eqn:normal}. This system is far larger than is convenient to solve
by any variant of Gaussian Elimination, even for our small 2D
examples, so iterative methods are 
necessary. We choose the Conjugate Gradient (CG.) method
\cite[]{Dan:67,Steihaug:83,NocedalWright}, and stop the
iteration by monitoring the reduction in length of the normal residual
(the gradient of $J_{\alpha,\sigma}[m,u;d]$ with respect to $u$). The
reduction threshhold $\rho$ is thus another necessary input
parameter. For all calculations in this paper, we use
we choose $\rho = 0.01$ (note that unlike $\alpha$ and $\sigma$,
$\rho$ is dimensionless).

Choice of regularization weights like $\alpha$ and $\sigma$ is a
widely studied topic; we mention some methods for this task in the
discussion section. For the set of examples presented here, we take a
simpler approach. Since the purpose of $\sigma>0$ is to avoid singularity
in the system \ref{eqn:normal} even for $\alpha=0$, we choose it a bit
bigger than floating point tolerance for single precision:
$\sigma = 10^{-5}$. We choose $\alpha$ by computing the solution
$u_{\alpha,\sigma}[m_0;d]$ for the data $d$ used in the first example
below, and several powers of 10 for $\alpha$. The largest such
$\alpha$ that yields an RMS predicted data error
($\|S[m_0]u_{\alpha,\sigma}[m_0;d] - d\|$) less than $0.05 \|d\|$ is
$\alpha = 10^{-4}$. We take this relation as representative, and use
this value for all MSWI computations.

\subsection{Recovering a circular lens}

The target bulk modulus field ($\kappa$) for the first example 
is depicted in Figure \ref{fig:m}. It contains an acoustic ``lens''
positioned in the center between 1000 m and 3000 m depth
(``$z$''). The background level outside the lens is 4 GPa; at the
center it is 2.4 GPa. The
buoyancy ($\beta$)is spatially homogeneous at 1 cm$^3/g$, so the background
wave speed is 2000 m/s.

@@

Simulated data from this configuration appears as Figure
\ref{fig:cwd20}.

\plot{cwd20}{width=\textwidth}{Data for circular lens model (\ref{fig:m}): 20
  sources at $x=3000$ m, 201 receivers at $x=5000$ m.}

The value of $J_{\rm FWI}[m_0,d] \approx 4.6$, and the rate of
increase in the weighted gradient direction at $m_0$ is 4.2 $\times
10^{-6 }$. Note that the rate of increase is the same as the weighted
gradient norm. After 12 steepest descent steps, the objective value has
decreased to 2.6, and the
weighted gradient norm by more than an order of magnitude, to $\approx 2.5 \times 10^{-7}$. The
final model appears in Figure \ref{fig:cwlens20mestfwi0}, and the
corresponding data in Figure \ref{fig:cwlens20resimmestfwi0}.

@@

\plot{cwlens20mestfwi0}{width=\textwidth}{a: Bulk modulus produced by 10
  steepest descent steps to minimize the FWI objective \ref{eqn:fwi},
  using the data shown in \ref{fig:cwd20}.}

\multiplot{2}{cwlens20resimmestfwi0,cwlens20residmestfwi0}{width=0.45\textwidth}{a: Data
  corresponding to FWI inversion result shown in Figure
  \ref{fig:cwlens20mestfwi0}. b: Residual or data error - difference between
  data shown in Figure \ref{fig:cwlens20resimmestfwi0} and target data
  \ref{fig:cwd20}. Note the failure to match arrival times of the later signal in the
  central part of the display.}

@@

This example is the one on which we based the choice $\alpha =
10^{-4}$. We did this by minimizing $J_{\alpha,\sigma}[m_0, u; d]$
over $u$ to identify a filter that outputs a good fit to the data
\ref{fig:cwd20} from the input \ref{fig:chwd20} (simulated from
$m_0$), and choosing $\alpha$ just small enough to obtain an RMS error
under 5\%.  The resulting adaptive filter $u_{\alpha,\sigma}[m_0;d]$
is shown in Figure \ref{fig:cwlens20uest0}. This filter exhibits a lot
of energy at non-zero times, as is necessary to move the events in
\ref{fig:chwd20} near those in \ref{fig:cwd20}. Thus the penalty term
is rather large. We will use the apparent dispersion of energy away
from $t=0$ in these adaptive filters to judge the quality of
inversions.

\plot{cwlens20uest0}{width=\textwidth}{Adaptive filter to match
  initial model data (Figure \ref{fig:chwd20}) to circular lens
  data (Figure \ref{fig:cwd20}). Note considerable energy dispersion
  away from $t=0$.}

We applied the same steepest descent algorithm to the MSWI objective
$J_{\alpha,\sigma}$. The initial bulk modulus field is once again
homogeneous at 4 GPa. The initial value of the reduced MSWI objective
is $\approx 1.49 \times 10^{-2}$, and the length of the (weighted)
gradient is $\approx 4.7 \times 10^{-10}$. After 12 iterations, the value has
decreased to $\approx 3.25 \times 10^{-3}$, or roughly a factor of 4.5.
reduction. The gradient length is
$\approx 5.6 \times 10^{-12}$, a reduction of almost two orders of
magnitude. The resulting bulk modulus is shown in Figure
\ref{fig:cwlens20mestmswi}.

\plot{cwlens20mestmswi}{width=\textwidth}{a: Bulk modulus produced by 12
  steepest descent steps to minimize the reduced MSWI objective \ref{eqn:redfiltpen},
  using the data shown in \ref{fig:cwd20}.}

The final adaptive filter is shown in Figure
\ref{fig:cwlens20uestmswi}. Note that energy in the filter has
migrated towards $t=0$, compared to the adaptive filter produced from
the homogenous bulk modulus \ref{fig:cwlens20uest0}, and the bulk of
it is within an apparent half-wavelength of $t=0$. This suggests that
the MSWI result may be an adequate initial estimate for a successful
FWI.

\plot{cwlens20uestmswi}{width=\textwidth}{Adaptive filter to match
  data for MSWI model (Figure \ref{fig:cwlens20resimmestmswi}) to circular lens
  data (Figure \ref{fig:cwd20}). Note considerable less energy dispersion
  away from $t=0$ than is evident in Figure \ref{fig:cwlens20uest0}.}

Another way to look at the improved data kinematics of the MSWI result
is via the simulate data from the model shown in
\ref{fig:cwlens20mestmswi}. Figures \ref{fig:cwlens20resimmestmswi},
\ref{fig:cwlens20residmestmswi} suggest that the
data predicted from the MSWI result is in many places within a
half-wavelength travel time shift of the data in Figure \ref{fig:cwd20}.
A half-wavelength is the conventional threshhold for FWI initial fit
error: below that it works, above that it doesn't.

\multiplot{2}{cwlens20resimmestmswi,cwlens20residmestmswi}{width=0.45\textwidth}{a:
  Data corresponding to MSWI result shown in Figure
  \ref{fig:cwlens20mestmswi}. b: Residual or data error - difference between
  data shown in Figure \ref{fig:cwlens20resimmestmswi} and target data
  \ref{fig:cwd20}. Note that the later signal in the central portion is
  still shifted, but by much less than in Figure \ref{fig:cwlens20residmestfwi0}.}

We applied 11 iterations of steepest descent to the FWI objective,
with initial estimate of $m$ equal to the final estimate from the MSWI
inversion just described (Figure \ref{fig:cwlens20mestmswi}). The initial FWI
objective value was $\approx 2.9$, and the weighted gradient norm was
$\approx 2.0 \times 10^{-4}$.  After 11 steps of steepest descent,
the prescribed reduction ($10^{-2}$) of the weighted gradient norm was achieved,
and the objective value had decreased to $\approx 9.1 \times
10^{-2}$. The resulting bulk modulus field is depicted in Figure
\ref{fig:cwlens20mestmswifwi}. The predicted data at this model, and the residual
data, are shown in Figures \ref{fig:cwlens20resimmestmswifwi} and \ref{fig:cwlens20residmestmswifwi}
respectively.
